---
title: "How to manage users, tokens and service accounts"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
```



For most of your workflow, you would follow a typical pattern:

1. Create a database, schema or table
2. Upload, update or amend data to your table
3. Query the data to do calculations, analytics or reference in dashboards

If you're running a database that multiple users have access to (either through cloud or on a local server) then a typical workflow may be

1. Create a users with certain set of permissions and roles to a database
2. Assign tokens / secrets to users so that they can access databaes remotely
3. Manage database utilities such as scaling
4. Run analytics on your database metadata

Let's see these in action

::: {.callout-note collapse="true"}
## DBI vs. md

These functions are mainly sugar syntax wrappers around the fabulous DBI packages to stream and simplify common database queries.

:::

# Database adminstrative functions
- list_extensions()
- install_extensions()
- load_extensions
- validate_install_status
- validate_load_status
- show_duckdb_settings()

-   These are a collection of motherduck specific database utilities to help you list, install and load duckdb extensions
-   the full list of extensions is available via duckdb community store are listed [here](https://duckdb.org/docs/stable/core_extensions/overview.html)
-   the `list_extensions()` will list duckdb extension from the community store and list their status, install or loaded
-   you can use `install_extensions()` to install a new exnsion -- this will also automatically load the extension
-   If the extension is already install you can use xx


```{r}
#| label: meta-example1
#| echo: true
#| eval: false
#| warning: false
#| message: false
#| include: true
list_extensions(con)

```


